draft

### 2025-05-23 draft.tex
Updates from Overleaf

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset. Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features refer to the nucleus and intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figs/spatial.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}
\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they can present barriers for smooth adoption into computational workflows. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). Although there are still rough edges to polish, such as ensuring a 1:1 match, it enables simpler automated analyses of microscopy data in either short scripts and complex pipelines. It avoids graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology. When used in tandem with generalist image-processing tools, such as Cellpose for segmentation (CITE Cellpose), we can streamline machine and deep learning workflows.

Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. We therefore foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits: This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the original community for which it was originally developed.

We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking. A functioning test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity. Once tests are in place, we could add to support just-in-time compiling and GPUs. Finally, we envision the cp\_measure library could be the place to develop and distribute new measurements from the community.

@@

This work was supported by GSK and the National Institutes of Health, grant (R35 GM122547 to AEC). The authors would like to thank Nodar Gogoberidze for his valuable insight into CellProfiler's code base, and Minh Doan and Eliot McKinley for their valuable feedback and discussions during the course of this project.
\end{document}

>>>>

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset. Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts \cite{garcia-fossaInterpretingImagebasedProfiles2023}.

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte nuclei image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features refer to the nucleus and intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figs/spatial.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}
\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they can present barriers for smooth adoption into computational workflows. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). Although there are still rough edges to polish, such as ensuring features match 1:1, it enables simpler automated profiling of microscopy data in short scripts and complex pipelines. It avoids graphical interfaces, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology. When used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.

Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. We therefore foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits. It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.

We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking. A functioning test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity. Once tests are in place, we could add support for just-in-time compiling and GPUs. We envision the cp\_measure library could also be the place to develop and distribute new measurements from the community.

@@

This work was supported by GSK and the National Institutes of Health, grant (R35 GM122547 to AEC). The authors would like to thank Nodar Gogoberidze, Beth Cimini, Minh Doan and Eliot McKinley for their valuable feedback and discussions during the course of this project.
\end{document}
